---
title: "2.4 — Costs of Production — Class Content"
draft: false
linktitle: "2.4 — Costs of Production"
date: "2020-06-08"
menu:
  content:
    parent: Course content
    weight: 11
type: docs
output:
  blogdown::html_page:
    toc: true
slides: "2.4-slides"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=F,
                      message=F,
                      warning=F,
                      fig.retina = 3,
                      fig.align="center")
library("tidyverse")
library("ggrepel")
update_geom_defaults("label", list(family = "Fira Sans Condensed"))
update_geom_defaults("text", list(family = "Fira Sans Condensed"))
```

{{% alert note %}}
*Wednesday, October 13, 2021*
{{% /alert %}}

{{% alert warning %}}
[Problem Set 3](/assignments/03-problem-set) is due by the end of the day Monday October 18.
{{% /alert %}}

## <i class="fas fa-info-circle fa-lg"></i> Overview

Today we cover costs before we put them together next class with revenues to solve the firm’s (first stage) *profit maximization problem*. While it seems we are adding quite a bit, and learning some new math problems with calculating costs, we will practice it more next class, when put together with revenues.

## <i class="fas fa-book-reader fa-lg"></i> Readings

- <i class="fas fa-book"></i> Ch. 7 in Goolsbee, Levitt, and Syverson, 2019

## <i class="fas fa-chalkboard-teacher"></i> Slides

Below, you can find the slides in two formats. Clicking the image will bring you to the html version of the slides in a new tab. Note while in going through the slides, you can type <kbd>h</kbd> to see a special list of viewing options, and type <kbd>o</kbd> for an outline view of all the slides.

The lower button will allow you to download a PDF version of the slides. I suggest printing the slides beforehand and using them to take additional notes in class (*not everything* is in the slides)!

`r blogdown::shortcode("slide-links")`

## <i class="fas fa-laptop-code"></i> Assignments

## Problem Set 3 Due Mon Oct 18

[Problem set 3](/assignments/03-problem-set) (on classes 2.1-2.3) is due by the end of the day Monday, October 18 by upload to Blackboard Assignments.

## <i class="far fa-sigma"></i> Appendix

### The Relationship Between Returns to Scale and Costs

There is a direct relationship between a technology's returns to scale^[Increasing, decreasing, or constant] and its cost structure: the rate at which its total costs increase^[At a decreasing rate, at an increasing rate, or at a constant rate, respectively] and its marginal costs change^[Decreasing, increasing, or constant, respectively]. This is easiest to see for a single input, such as our assumptions of the short run (where firms can change $l$ but not $\bar{k})$:

$$q=f(\bar{k},l)$$

### Constant Returns to Scale: 

```{r, fig.retina=3, fig.height=5.5}
crs_prod=function(x){x}

crs_q<-ggplot(data.frame(x=c(0,10)), aes(x=x))+
  stat_function(fun=crs_prod, geom="line", size=2, color="red")+
  geom_label(aes(x=2.5,y=crs_prod(2.5)), color = "red", label="Total Product", size = 5)+
  # first dotted lines
  geom_segment(x=1, xend=1, y=0, yend=1, linetype="dotted",size=1)+
  geom_segment(x=0, xend=1, y=1, yend=1, linetype="dotted",size=1)+
  # second dotted lines
  geom_segment(x=2, xend=2, y=0, yend=2, linetype="dotted",size=1)+
  geom_segment(x=0, xend=2, y=2, yend=2, linetype="dotted",size=1)+
    scale_x_continuous(breaks=c(0,1,2),
                       labels=c(0,"l","2l"),
                     limits=c(0,3),
                     expand=expand_scale(mult=c(0,0.1)))+
  scale_y_continuous(breaks=c(0,1,2),
                     labels=c(0,"f(l)", "f(2l)"),
                     limits=c(0,3),
                     expand=expand_scale(mult=c(0,0.1)))+
  guides(color = F)+
  labs(x = "Input, Labor (l)",
       y = "Output (q)",
       title = "Constant Returns")+
  theme_classic(base_family = "Fira Sans Condensed", base_size=14)

crs_cost=function(x){x}

crs_tc<-ggplot(data.frame(x=c(0,10)), aes(x=x))+
  stat_function(fun=crs_cost, geom="line", size=2, color="red")+
  #geom_label(aes(x=2.5,y=crs_prod(2.5)), color = "red", label="Total Product", size = 5)+
    scale_x_continuous(breaks=c(0),
                       labels=c(0),
                     limits=c(0,3),
                     expand=expand_scale(mult=c(0,0.1)))+
  scale_y_continuous(breaks=c(0),
                     labels=c(0),
                     limits=c(0,3),
                     expand=expand_scale(mult=c(0,0.1)))+
  guides(color = F)+
  labs(x = "Output (q)",
       y = "Total Costs C(q)",
       title = "Total Costs")+
  theme_classic(base_family = "Fira Sans Condensed", base_size=14)

crs_mc=function(x){2}

crs_mc<-ggplot(data.frame(x=c(0,10)), aes(x=x))+
  stat_function(fun=crs_mc, geom="line", size=2, color="red")+
  geom_label(aes(x=2.5,y=2), color = "red", label="MC(q)=AC(q)", size = 5)+
    scale_x_continuous(breaks=c(0),
                       labels=c(0),
                     limits=c(0,3),
                     expand=expand_scale(mult=c(0,0.1)))+
  scale_y_continuous(breaks=c(0),
                     labels=c(0),
                     limits=c(0,3),
                     expand=expand_scale(mult=c(0,0.1)))+
  guides(color = F)+
  labs(x = "Output (q)",
       y ="Costs per Unit",
       title = "Per Unit Costs")+
  theme_classic(base_family = "Fira Sans Condensed", base_size=14)

library(patchwork)
crs_q+(crs_tc/crs_mc)
```

### Decreasing Returns to Scale

```{r, fig.retina=3, fig.height=5.5}
library(mosaic)
drs_prod=function(x){sqrt(x)}

drs_q<-ggplot(data.frame(x=c(0,10)), aes(x=x))+
  stat_function(fun=drs_prod, geom="line", size=2, color="red")+
  geom_abline(intercept=0, slope=1, size=1, linetype="dashed")+
  geom_label(aes(x=2.5,y=drs_prod(2.5)), color = "red", label="Total Product", size = 5)+
  geom_label(aes(x=3,y=3), label=expression(45*degree), size = 5)+
  # first dotted lines
  geom_segment(x=1, xend=1, y=0, yend=1, linetype="dotted",size=1)+
  geom_segment(x=0, xend=1, y=1, yend=1, linetype="dotted",size=1)+
  # second dotted lines
  geom_segment(x=2, xend=2, y=0, yend=2, linetype="dotted",size=1)+
  geom_segment(x=0, xend=2, y=2, yend=2, linetype="dotted",size=1)+
  # function dotted lines
  geom_segment(x=2, xend=2, y=0, yend=drs_prod(2), linetype="dotted",size=1)+
  geom_segment(x=0, xend=2, y=drs_prod(2), yend=drs_prod(2), linetype="dotted",size=1)+
    scale_x_continuous(breaks=c(0,1,2),
                       labels=c(0,"l","2l"),
                     limits=c(0,3),
                     expand=expand_scale(mult=c(0,0.1)))+
  scale_y_continuous(breaks=c(0,1,drs_prod(2),2),
                     labels=c(0,"f(l)", "2f(l)", "f(2l)"),
                     limits=c(0,3),
                     expand=expand_scale(mult=c(0,0.1)))+
  guides(color = F)+
  labs(x = "Input, Labor (l)",
       y = "Output (q)",
       title = "Decreasing Returns")+
  theme_classic(base_family = "Fira Sans Condensed", base_size=14)

drs_cost=function(x){x^{1.5}}

drs_tc<-ggplot(data.frame(x=c(0,10)), aes(x=x))+
  stat_function(fun=drs_cost, geom="line", size=2, color="red")+
  #geom_label(aes(x=2.5,y=crs_prod(2.5)), color = "red", label="Total Product", size = 5)+
    scale_x_continuous(breaks=c(0),
                       labels=c(0),
                     limits=c(0,3),
                     expand=expand_scale(mult=c(0,0.1)))+
  scale_y_continuous(breaks=c(0),
                     labels=c(0),
                     limits=c(0,3),
                     expand=expand_scale(mult=c(0,0.1)))+
  guides(color = F)+
  labs(x = "Output (q)",
       y = "Total Costs C(q)",
       title = "Total Costs")+
  theme_classic(base_family = "Fira Sans Condensed", base_size=14)

drs_mc=function(x){x^{1.5}}

drs_mc2<-ggplot(data.frame(x=c(0,10)), aes(x=x))+
  stat_function(fun=drs_mc, geom="line", size=2, color="red")+
  geom_label(aes(x=2,y=drs_mc(2)), color = "red", label="MC(q)", size = 5)+
    scale_x_continuous(breaks=c(0),
                       labels=c(0),
                     limits=c(0,3),
                     expand=expand_scale(mult=c(0,0.1)))+
  scale_y_continuous(breaks=c(0),
                     labels=c(0),
                     limits=c(0,3),
                     expand=expand_scale(mult=c(0,0.1)))+
  guides(color = F)+
  labs(x = "Output (q)",
       y ="Costs per Unit",
       title = "Per Unit Costs")+
  theme_classic(base_family = "Fira Sans Condensed", base_size=14)

drs_q+(drs_tc/drs_mc2)
```

### Increasing Returns to Scale

```{r, fig.retina=3, fig.height=5.5}
library(mosaic)
irs_prod=function(x){x^{1.5}}

irs_q<-ggplot(data.frame(x=c(0,10)), aes(x=x))+
  stat_function(fun=irs_prod, geom="line", size=2, color="red")+
  geom_abline(intercept=0, slope=1, size=1, linetype="dashed")+
  geom_label(aes(x=2.5,y=irs_prod(2.5)), color = "red", label="Total Product", size = 5)+
  geom_label(aes(x=3,y=3), label=expression(45*degree), size = 5)+
  # first dotted lines
  geom_segment(x=1, xend=1, y=0, yend=1, linetype="dotted",size=1)+
  geom_segment(x=0, xend=1, y=1, yend=1, linetype="dotted",size=1)+
  # second dotted lines
  geom_segment(x=2, xend=2, y=0, yend=2, linetype="dotted",size=1)+
  geom_segment(x=0, xend=2, y=2, yend=2, linetype="dotted",size=1)+
  # function dotted lines
  geom_segment(x=2, xend=2, y=0, yend=irs_prod(2), linetype="dotted",size=1)+
  geom_segment(x=0, xend=2, y=irs_prod(2), yend=irs_prod(2), linetype="dotted",size=1)+
    scale_x_continuous(breaks=c(0,1,2),
                       labels=c(0,"l","2l"),
                     limits=c(0,3),
                     expand=expand_scale(mult=c(0,0.1)))+
  scale_y_continuous(breaks=c(0,1,2, irs_prod(2)),
                     labels=c(0,"f(l)", "2f(l)", "f(2l)"),
                     limits=c(0,3),
                     expand=expand_scale(mult=c(0,0.1)))+
  guides(color = F)+
  labs(x = "Input, Labor (l)",
       y = "Output (q)",
       title = "Increasing Returns")+
  theme_classic(base_family = "Fira Sans Condensed", base_size=14)

irs_cost=function(x){x^{0.5}}

irs_tc<-ggplot(data.frame(x=c(0,10)), aes(x=x))+
  stat_function(fun=irs_cost, geom="line", size=2, color="red")+
  #geom_label(aes(x=2.5,y=crs_prod(2.5)), color = "red", label="Total Product", size = 5)+
    scale_x_continuous(breaks=c(0),
                       labels=c(0),
                     limits=c(0,3),
                     expand=expand_scale(mult=c(0,0.1)))+
  scale_y_continuous(breaks=c(0),
                     labels=c(0),
                     limits=c(0,3),
                     expand=expand_scale(mult=c(0,0.1)))+
  guides(color = F)+
  labs(x = "Output (q)",
       y = "Total Costs C(q)",
       title = "Total Costs with Increasing RTS")+
  theme_classic(base_family = "Fira Sans Condensed", base_size=14)

irs_mc=function(x){1/x}

irs_mc2<-ggplot(data.frame(x=c(0,10)), aes(x=x))+
  stat_function(fun=irs_mc, geom="line", size=2, color="red")+
  geom_label(aes(x=2.5,y=irs_mc(2.5)), color = "red", label="MC(q)", size = 5)+
    scale_x_continuous(breaks=c(0),
                       labels=c(0),
                     limits=c(0,3),
                     expand=expand_scale(mult=c(0,0.1)))+
  scale_y_continuous(breaks=c(0),
                     labels=c(0),
                     limits=c(0,3),
                     expand=expand_scale(mult=c(0,0.1)))+
  guides(color = F)+
  labs(x = "Output (q)",
       y ="Costs per Unit",
       title = "Per Unit Costs")+
  theme_classic(base_family = "Fira Sans Condensed", base_size=14)

irs_q+(irs_tc/irs_mc2)
```


### Cobb-Douglas Cost Functions

The total cost function for Cobb-Douglas production functions of the form
$$q=l^{\alpha}k^{\beta}$$
can be shown with some *very tedious* algebra to be:

$$C(w,r,q)=\left[\left(\frac{\alpha}{\beta}\right)^{\frac{\beta}{\alpha+\beta}} + \left(\frac{\alpha}{\beta}\right)^{\frac{-\alpha}{\alpha+\beta}}\right] w^{\frac{\alpha}{\alpha+\beta}} r^{\frac{\beta}{\alpha+\beta}} q^{\frac{1}{\alpha+\beta}}$$

If you take the first derivative of this (to get marginal cost), it is:

$$\frac{\partial C(w,r,q)}{\partial q}= MC(q) = \frac{1}{\alpha+\beta} \left(w^{\frac{\alpha}{\alpha+\beta}} r^{\frac{\beta}{\alpha+\beta}}\right) q^{\left(\frac{1}{\alpha+\beta}\right)-1}$$

How does marginal cost change with increased output? Take the second derivative:

$$\frac{\partial^2 C(w,r,q)}{\partial q^2}= \frac{1}{\alpha+\beta} \left(\frac{1}{\alpha+\beta} -1 \right) \left(w^{\frac{\alpha}{\alpha+\beta}} r^{\frac{\beta}{\alpha+\beta}}\right) q^{\left(\frac{1}{\alpha+\beta}\right)-2}$$

Three possible cases:

1. If $\frac{1}{\alpha+\beta} > 1$, this is positive $\implies$ **_decreasing_ returns to scale**
  - Production function exponents $\alpha+\beta < 1$
2. If $\frac{1}{\alpha+\beta} < 1$, this is negative $\implies$ **_increasing_ returns to scale**
  - Production function exponents $\alpha+\beta > 1$
3. If $\frac{1}{\alpha+\beta} = 1$, this is constant $\implies$ **_constant_ returns to scale**
  - Production function exponents$\alpha+\beta = 1$

#### Example (Constant Returns)

Let $q=l^{0.5}k^{0.5}$. 

$$\begin{align*}
C(w,r,q)&=\left[\left(\frac{0.5}{0.5}\right)^{\frac{0.5}{0.5+0.5}} + \left(\frac{0.5}{0.5}\right)^{\frac{-0.5}{0.5+0.5}}\right] w^{\frac{0.5}{0.5+0.5}} r^{\frac{0.5}{0.5+0.5}} q^{\frac{1}{0.5+0.5}}\\
C(w,r,q)&= \left[1^{0.5}+1^{-0.5} \right]w^{0.5}r^{0.5}q^{0.5}\\
C(w,r,q)&= w^{0.5}r^{0.5}q^{1}\\
\end{align*}$$

Consider input prices of $w=\$9$ and $r=\$25$: 

$$\begin{align*}C(w=9,r=25,q)&=9^{0.5}25^{0.5}q \\ 
& =3*5*q\\
& =15q\\\end{align*}$$

That is, total costs (at those given input prices, and technology) is equal to 15 times the output level, $q$:

```{r, fig.retina=3, fig.height=5.5}
cost_ex=function(x){15*x}

ggplot(data.frame(x=c(0,10)), aes(x=x))+
  stat_function(fun=cost_ex, geom="line", size=2, color="red")+
  geom_label(aes(x=19,y=cost_ex(19)), color = "red", label="C(q) = 15q", size = 5)+
    scale_x_continuous(breaks=seq(0,20,2),
                     limits=c(0,20),
                     expand=expand_scale(mult=c(0,0.1)))+
  scale_y_continuous(breaks=seq(0,400,50),
                     limits=c(0,400),
                     expand=expand_scale(mult=c(0,0.1)))+
  guides(color = F)+
  labs(x = "Output (q)",
       y ="Costs C(q)",
       title = "Total Costs")+
  theme_classic(base_family = "Fira Sans Condensed", base_size=16)
```

Marginal costs would be

$$MC(q) = \frac{\partial C(q)}{\partial q} = 15$$

Average costs would be

$$MC(q) = \frac{C(q)}{q} = \frac{15q}{q} = 15$$

```{r, fig.retina=3, fig.height=5.5}
mc_ex=function(x){15}

ggplot(data.frame(x=c(0,10)), aes(x=x))+
  stat_function(fun=mc_ex, geom="line", size=2, color="red")+
  geom_label(aes(x=17,y=15), color = "red", label="MC(q)=AC(q)", size = 5)+
    scale_x_continuous(breaks=seq(0,20,2),
                     limits=c(0,20),
                     expand=expand_scale(mult=c(0,0.1)))+
  scale_y_continuous(breaks=seq(0,20,2),
                     limits=c(0,20),
                     expand=expand_scale(mult=c(0,0.1)))+
  guides(color = F)+
  labs(x = "Output (q)",
       y ="Per Unit Costs",
       title = "Per Unit Costs")+
  theme_classic(base_family = "Fira Sans Condensed", base_size=16)
```

#### Example (Decreasing Returns)

Let $q=l^{0.25}k^{0.25}$. 

$$\begin{align*}
C(w,r,q)&=\left[\left(\frac{0.25}{0.25}\right)^{\frac{0.25}{0.25+0.25}} + \left(\frac{0.25}{0.25}\right)^{\frac{-0.25}{0.25+0.25}}\right] w^{\frac{0.25}{0.25+0.25}} r^{\frac{0.25}{0.25+0.25}} q^{\frac{1}{0.25+0.25}}\\
C(w,r,q)&= \left[1^{0.5}+1^{-0.5} \right]w^{0.5}r^{0.5}q^{2}\\
C(w,r,q)&= w^{0.5}r^{0.5}q^{2}\\
\end{align*}$$

If $w=9$, $r=25$: 

$$\begin{align*}C(w=9,r=25,q)&=9^{0.5}25^{0.5}q^2 \\ 
& =3*5*q^2\\
& =15q^2\\\end{align*}$$

```{r, fig.retina=3, fig.height=5.5}
cost_ex=function(x){15*x^2}

ggplot(data.frame(x=c(0,10)), aes(x=x))+
  stat_function(fun=cost_ex, geom="line", size=2, color="red")+
  geom_label(aes(x=16,y=cost_ex(16)), color = "red", label="C(q)", size = 5)+
    scale_x_continuous(breaks=seq(0,20,2),
                     limits=c(0,20),
                     expand=expand_scale(mult=c(0,0.1)))+
  scale_y_continuous(breaks=seq(0,5000,500),
                     limits=c(0,5000),
                     expand=expand_scale(mult=c(0,0.1)))+
  guides(color = F)+
  labs(x = "Output (q)",
       y ="Costs C(q)",
       title = "Total Costs")+
  theme_classic(base_family = "Fira Sans Condensed", base_size=16)
```

Marginal costs would be

$$MC(q) = \frac{\partial C(q)}{\partial q} = 30q$$

Average costs would be

$$AC(q) = \frac{C(q)}{q} = \frac{15q^2}{q} = 15q$$

```{r, fig.retina=3, fig.height=5.5}
mc_ex_2=function(x){30*x}
ac_ex_2=function(x){15*x}

ggplot(data.frame(x=c(0,10)), aes(x=x))+
  stat_function(fun=mc_ex_2, geom="line", xlim=c(0,20), size=2, color="red")+
  stat_function(fun=ac_ex_2, geom="line", xlim=c(0,20), size=2, color="orange")+
  geom_label(aes(x=8,y=mc_ex_2(8)), color = "red", label="MC(q)", size = 5)+
  geom_label(aes(x=18,y=ac_ex_2(18)), color = "orange", label="AC(q)", size = 5)+
    scale_x_continuous(breaks=seq(0,20,2),
                     limits=c(0,20),
                     expand=expand_scale(mult=c(0,0.1)))+
  scale_y_continuous(breaks=seq(0,300,50),
                     limits=c(0,300),
                     expand=expand_scale(mult=c(0,0.1)))+
  guides(color = F)+
  labs(x = "Output (q)",
       y ="Per Unit Costs",
       title = "Per Unit Costs")+
  theme_classic(base_family = "Fira Sans Condensed", base_size=16)
```