---
title: "I & S Effects: Price *Increase*"
subtitle: "ECON 306 • Microeconomic Analysis • Spring 2021"
author: 'Ryan Safner<br> Assistant Professor of Economics <br> <a href="mailto:safner@hood.edu"><i class="fa fa-paper-plane fa-fw"></i>safner@hood.edu</a> <br> <a href="https://github.com/ryansafner/microS21"><i class="fa fa-github fa-fw"></i>ryansafner/microS21</a><br> <a href="https://microS21.classes.ryansafner.com"> <i class="fa fa-globe fa-fw"></i>microS21.classes.ryansafner.com</a><br>'
#date:
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML" # rescales math with css changes https://github.com/yihui/xaringan/issues/143
    lib_dir: libs
    #seal: false
    css: [custom.css, "hygge"] #, metropolis, metropolis-fonts
    nature:
      beforeInit: ["macros.js", "https://platform.twitter.com/widgets.js"] # first is for rescaling images , second is for embedding tweets, https://github.com/yihui/xaringan/issues/100
      highlightStyle: ocean #rainbow #monokai rainbow solarized-light #tomorrow-night -blue -eighties zenburn
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
          </div>
        </div>
    includes:
      in_header: header.html # for font awesome, used in title  
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo=F,
                      message=F,
                      warning=F,
                      fig.retina = 3,
                      fig.align = "center")
library("tidyverse")
library("mosaic")
library("ggrepel")
library("fontawesome")
xaringanExtra::use_tile_view()
xaringanExtra::use_tachyons()

update_geom_defaults("label", list(family = "Fira Sans Condensed"))
update_geom_defaults("text", list(family = "Fira Sans Condensed"))

set.seed(256)
```

# Real Income and Substitution Effects, Graphically I

.pull-left[

- Original optimal consumption $(A)$

]

.pull-right[

```{r}
bc<-function(x){24-x}
ic<-function(x){144/x}
points<-tribble(
  ~x, ~y, ~letter,
  12, 12, "A"
)
opt<-ggplot(data = tibble(x = c(0,10)), aes(x = x))+
  geom_segment(x = 12, xend = 12, y = 0, yend = 12, linetype = "dashed", size = 1)+
  geom_segment(x = 0, xend = 12, y = 12, yend = 12, linetype = "dashed", size = 1)+
  geom_function(fun = bc, geom = "line", size = 2, color = "red")+
  geom_label(x = 2, y = bc(2), color = "red", label = "BC")+
  geom_function(fun = ic, geom = "line", size = 2, color = "blue")+
  geom_label(x = 22, y = ic(22), color = "blue", label = expression(u[1]))+
  geom_point(data = subset(points, letter %in% c("A")), aes(x = x, y = y), size = 5)+
  scale_x_continuous(breaks = seq(0,24,2),
  #                   labels = c(0, expression("\u2113"^"*"), 24),
                     limits = c(0,25),
                     expand = c(0,0))+
  scale_y_continuous(breaks = seq(0,24,2),
                     #labels = c(0,expression(c^"*"),"24w"),
                     expand = c(0,0),
                     limits = c(0,25))+
  labs(x = "x",
       y = "y",
       caption = expression(paste("Optima with ", u(x,y)==x^{0.5}*y^{0.5}, ", ", m==24, ", ", p[y]==1, sep=" ")))+
  theme_classic(base_family = "Fira Sans Condensed", base_size = 16)
opt
```
]

---

# Real Income and Substitution Effects, Graphically I

.pull-left[

- Original optimal consumption $(A)$

- .purple[**(Total) price effect:** `\\(A \rightarrow C\\)`]

- Let's decompose this into the two effects

]

.pull-right[
```{r}
bc2<-function(x){24-2*x}
ic<-function(x){144/x}
points<-tribble(
  ~x, ~y, ~letter,
  12, 12, "A",
  9, 8, "B",
  6, 12, "C"
)
ic2<-function(x){72/x}
bc_sub<-function(x){34-2*x}
opt2<-ggplot(data = tibble(x = c(0,10)), aes(x = x))+
  geom_segment(x = 12, xend = 12, y = 0, yend = 12, linetype = "dashed", size = 0.5)+
  geom_segment(x = 0, xend = 12, y = 12, yend = 12, linetype = "dashed", size = 0.5)+
  
  geom_segment(x = 6, xend = 6, y = 0, yend = 12, linetype = "dashed", size = 0.5)+
  geom_segment(x = 0, xend = 6, y = 12, yend = 12, linetype = "dashed", size = 0.5)+
  geom_function(fun = bc, geom = "line", size = 2, color = "red")+
  geom_label(x = 2, y = bc(2), color = "red", label = expression(BC[1]))+
  geom_function(fun = ic, geom = "line", size = 2, color = "blue")+
  geom_label(x = 22, y = ic(22), color = "blue", label = expression(u[1]))+
  
  geom_function(fun = bc2, geom = "line", size = 2, color = "darkred")+
  geom_label(x = 2, y = bc2(2), color = "darkred", label = expression(BC[2]))+
  
  # geom_function(fun = bc_sub, geom = "line", size = 2, color = "orange", linetype = "dashed")+
  
    geom_function(fun = ic2, geom = "line", size = 2, color = "green")+
  geom_label(x = 22, y = ic2(22), color = "green", label = expression(u[2]))+
  geom_point(data = subset(points, letter %in% c("A", "C")), aes(x = x, y = y), size = 5)+
  ggrepel::geom_text_repel(data = subset(points, letter %in% c("A", "C")), aes(x = x, y = y, label = letter), size = 4, seed = 4, box.padding = 0.5)+
  annotate("segment", x = 12, xend = 6, y = 2, yend = 2, colour = "purple", size=2, alpha=1, arrow=arrow(length=unit(0.5,"cm"), ends="last", type="closed"))+
  scale_x_continuous(breaks = seq(0,24,2),
  #                   labels = c(0, expression("\u2113"^"*"), 24),
                     limits = c(0,25),
                     expand = c(0,0))+
  scale_y_continuous(breaks = seq(0,24,2),
                     #labels = c(0,expression(c^"*"),"24w"),
                     expand = c(0,0),
                     limits = c(0,25))+
  labs(x = "x",
       y = "y",
       caption = expression(paste("Optima with ", u(x,y)==x^{0.5}*y^{0.5}, ", ", m==24, ", ", p[y]==1, sep=" ")))+
  theme_classic(base_family = "Fira Sans Condensed", base_size = 16)
opt2
```

]

---

# Real Income and Substitution Effects, Graphically II

.pull-left[

- .orange[**Substitution effect:**] what you would choose under the **new exchange rate** to **remain indifferent** as before the change

]

.pull-right[

```{r, fig.retina=3, fig.height=6}
opt2
```

]

---

# Real Income and Substitution Effects, Graphically II

.pull-left[

- .orange[**Substitution effect:**] what you would choose under the **new exchange rate** to **remain indifferent** as before the change

- Graphically: shift *new* budget constraint inwards until tangent with *old* indifference curve

- $A \rightarrow B$ on same I.C. $(\uparrow$ $x$, $\downarrow$ $y)$
    - Point B *must* be a *different* point on the original curve! 

]

.pull-right[
```{r}
bc_sub<-function(x){17-x}
se_dom<-ggplot(data = tibble(x = c(0,10)), aes(x = x))+
  geom_function(fun = bc, geom = "line", size = 2, color = "red")+
  geom_label(x = 2, y = bc(2), color = "red", label = expression(BC[1]))+
  geom_function(fun = ic, geom = "line", size = 2, color = "blue")+
  geom_label(x = 22, y = ic(22), color = "blue", label = expression(u[1]))+
  
  geom_function(fun = bc2, geom = "line", size = 2, color = "darkred")+
  geom_label(x = 2, y = bc2(2), color = "darkred", label = expression(BC[2]))+
  
  
  geom_function(fun = bc_sub, geom = "line", size = 2, color = "orange", linetype = "dashed")+
  
  geom_function(fun = ic2, geom = "line", size = 2, color = "darkblue")+
  geom_label(x = 22, y = ic2(22), color = "darkblue", label = expression(u[0]))+
  geom_segment(x = 12, xend = 12, y = 0, yend = 12, linetype = "dashed", size = 0.5)+
  geom_segment(x = 0, xend = 12, y = 12, yend = 12, linetype = "dashed", size = 0.5)+
  
  geom_segment(x = 6, xend = 6, y = 0, yend = 12, linetype = "dashed", size = 0.5)+
  geom_segment(x = 0, xend = 6, y = 12, yend = 12, linetype = "dashed", size = 0.5)+
  
  geom_segment(x = 9, xend = 9, y = 0, yend = 8, linetype = "dashed", size = 0.5)+
  geom_segment(x = 0, xend = 9, y = 8, yend = 8, linetype = "dashed", size = 0.5)+
  geom_point(data = subset(points, letter %in% c("A", "B", "C")), aes(x = x, y = y), size = 5)+
  ggrepel::geom_text_repel(data = subset(points, letter %in% c("A","B", "C")), aes(x = x, y = y, label = letter), size = 4, seed = 4, box.padding = 0.5)+
  #annotate("segment", x = 12, xend = 10, y = 2, yend = 2, colour = "purple", size=2, alpha=1, arrow=arrow(length=unit(0.5,"cm"), ends="last", type="closed"))+
  annotate("segment", x = 12, xend = 9, y = 4, yend = 4, colour = "orange", size=2, alpha=1, arrow=arrow(length=unit(0.5,"cm"), ends="last", type="closed"))+
  scale_x_continuous(breaks = seq(0,24,2),
  #                   labels = c(0, expression("\u2113"^"*"), 24),
                     limits = c(0,25),
                     expand = c(0,0))+
  scale_y_continuous(breaks = seq(0,24,2),
                     #labels = c(0,expression(c^"*"),"24w"),
                     expand = c(0,0),
                     limits = c(0,25))+
  labs(x = "x",
       y = "y",
       caption = expression(paste("Optima with ", u(x,y)==x^{0.5}*y^{0.5}, ", ", m==24, ", ", p[y]==1, sep=" ")))+
  theme_classic(base_family = "Fira Sans Condensed", base_size = 16)
se_dom
```
]

---

# Real Income and Substitution Effects, Graphically III

.pull-left[

- .green[**(Real) income effect:**] change in consumption due to the **change in purchasing power** from the change in price

]

.pull-right[

```{r, fig.retina=3, fig.height=6}
se_dom
```

]

---

# Real Income and Substitution Effects, Graphically III

.pull-left[

- .green[**(Real) income effect:**] change in consumption due to the **change in purchasing power** from the change in price

- $B \rightarrow C$ to new budget constraint (can buy more of $x$ and/or $y$)
]

.pull-right[

```{r, fig.retina=3, fig.height=6}
se_dom+
  annotate("segment", x = 9, xend = 6, y = 4, yend = 4, colour = "green", size=2, alpha=1, arrow=arrow(length=unit(0.5,"cm"), ends="last", type="closed"))
```

]

---

# Real Income and Substitution Effects, Graphically IV

.pull-left[

- Original optimal consumption $(A)$

]

.pull-right[

```{r, fig.retina=3, fig.height=6}
opt
```

]

---

# Real Income and Substitution Effects, Graphically IV

.pull-left[

- Original optimal consumption $(A)$

- Price of $x$ falls, new optimal consumption at $(C)$

]

.pull-right[

```{r, fig.retina=3, fig.height=6}
opt2
```

]

---

# Real Income and Substitution Effects, Graphically IV

.pull-left[

- Original optimal consumption $(A)$

- Price of $x$ falls, new optimal consumption at $(C)$

- .orange[**Substitution effect:** `\\(A \rightarrow B\\)`] on same I.C. $(\uparrow$ cheaper $x$ and $\downarrow$ $y)$

]

.pull-right[

```{r, fig.retina=3, fig.height=6}
se_dom
```

]

---

# Real Income and Substitution Effects, Graphically IV

.pull-left[

- Original optimal consumption $(A)$

- Price of $x$ falls, new optimal consumption at $(C)$

- .orange[**Substitution effect:** `\\(A \rightarrow B\\)`] on same I.C. $(\uparrow$ cheaper $x$ and $\downarrow$ $y)$

- .green[**(Real) income effect:** `\\(B \rightarrow C\\)`] to new budget constraint (can buy more of $x$ and/or $y$)

]

.pull-right[

```{r, fig.retina=3, fig.height=6}
se_dom+
  annotate("segment", x = 8, xend = 10, y = 6, yend = 6, colour = "green", size=2, alpha=1, arrow=arrow(length=unit(0.5,"cm"), ends="last", type="closed"))
```

]

---

# Real Income and Substitution Effects, Graphically IV

.pull-left[

- Original optimal consumption $(A)$

- Price of $x$ falls, new optimal consumption at $(C)$

- .orange[**Substitution effect:** `\\(A \rightarrow B\\)`] on same I.C. $(\uparrow$ cheaper $x$ and $\downarrow$ $y)$

- .green[**(Real) income effect:** `\\(B \rightarrow C\\)`] to new budget constraint (can buy more of $x$ and/or $y$)

- .purple[**(Total) price effect:** `\\(A \rightarrow C\\)`]

]

.pull-right[

```{r, fig.retina=3, fig.height=6}
se_dom+
  annotate("segment", x = 9, xend = 6, y = 6, yend = 6, colour = "green", size=2, alpha=1, arrow=arrow(length=unit(0.5,"cm"), ends="last", type="closed"))+
    annotate("segment", x = 12, xend = 6, y = 2, yend = 2, colour = "purple", size=2, alpha=1, arrow=arrow(length=unit(0.5,"cm"), ends="last", type="closed"))+
  labs(title = "Change in Consumption From an Increase in Price",
       subtitle = "Normal Good")

```

]